Formal Paper Draft

### 2022-01-10 PaperDraft.Rmd
adding files

---
title: "Formal Paper Draft"
author: "Paul Harmon"
date: "8/21/2021"
output:
  pdf_document: default
  html_document: default
subtitle: Influence Diagnostics for High-Dimensional Ordination Techniques
---

@@

$$C = \sum_{i} \sum_{j} p_{j|i} log(\frac{p_{j|i}}{q_{j|i}})$$

t-SNE is an improvement over Stochastic Neightbor Embedding (SNE), which makes use of Gaussian distributions to model both $p_{ij}$ and $q_{ij}$ make preserving local structure somewhat difficult. Most of the minimization of objective functions is done via gradient descent (and some methods require simulated annealing because the algorithm can get caught in local optima). Critically, because the optimization is non-convex, and as a consequence of the gradient descent methods used to optimize it, t-SNE has the potential to provide different maps when run on the same data at the same perplexity.

@@

Additionally, t-SNE is subject to a pre-processing step involving principal component analysis (PCA).  In most implementations, "whitening" via PCA is performed on the initial dataset (or input distance matrix) to reduce the dimensionality to a reasonable value.  For instance, in Van der Maaten's initial paper, several examples reduced the initial dimensionality of the datato 30 for computational gains (Van der maaten, 2011).

@@

@@

Placeholder section in case we want to add another mapping technique in.

+ SNE
+ Sammon Mapping
+ Isomap

@@

In regression models, an outlier is typically denoted as a point that is either much larger or smaller than similar observations on a specific scale. However, an outlier need not be an influential point. An influential point simply needs to be outlying enough relative to similar observations that it impacts the fit of the regression line. the identification of influential points that can impact the results of a model is commonplace.

@@

Although Cook's distance is essentially an F statistic from an $F_{p, n-p}$ distribution, and the method does not use p-values to determine how "influential" a point is. Rather, cutoff values are typically assessed, or observations with relatively large Cook's distance values are considered for investigation of influence.

## Outliers and Influential Points in Maps

@@

@@

Second, distance matrices contain information on point-to-point distances (or dissimilarities) and, when vectorized, the correlation of the distance matrices measures similarity of two configurations (Mantel's test (Mantel, 1967) makes use of this approach to assess correlation between distance matrices).

Third, correlations (or similarities) can be converted to distances; one such metric for doing this is $\sqrt{2(1-r_{ij})}$ (James et al, 2013) where $r_{ij}$ is the correlation between $x_i$ and $x_j$.

@@

Our Permanova-based approach is as follows:

@@

We will discuss correcting p-values for multiple testing later, although currently our plan is to assess using Benjamini-Hochberg False Discovery Rate or Bonferroni corrections (or something similar, assuming that the results are based on p-values as in the Bonferroni Outlier test (Cook & Weisberg, 1982)). We would expect that if the maps are suitably distorted, or *influenced* by a single point, the resulting p-value for the jth adonis test would be small.

@@

```{r}
 g1 <- rmnorm(n = 10, mean = rep(20, 20), diag(35, 20)) %>%
    t() %>%
    data.frame()
  g2 <- rmnorm(n = 10, mean = rep(40, 20), diag(35, 20)) %>%
    t() %>%
    data.frame()
  g3 <- rmnorm(n = 10, mean = rep(60, 20), diag(35, 20)) %>%
    t() %>%
    data.frame()

@@

```{r}
#2021-10-18_simulationsmaller_mds_list

@@

#mean detection rates
Category <- c("Correctly Identified Influential", "Non-Influential")
TSNE <- c(round(mean(ci1),3), round(mean(cni1),3))
MDS <- c(round(mean(ci2),3), round(mean(cni2),3))
temp <- cbind(Category,TSNE, MDS)

kable(temp)

@@

#mean detection rates
Category <- c("Correctly Identified Influential", "Non-Influential")
TSNE <- c(mean(ci1), mean(cni1))
MDS <- c(mean(ci2), round(mean(cni2),3))
temp <- cbind(Category,TSNE, MDS)

kable(temp)

@@

#mean detection rates
Category <- c("Correctly Identified Influential", "Non-Influential")
TSNE <- c(mean(ci1), round(mean(cni1),3))
MDS <- c(mean(ci2), round(mean(cni2),3))
temp <- cbind(Category,TSNE, MDS)

kable(temp)

@@

In order to determine a notion of how severe an outlier/influential point might be, we investigated several methods for quantifying the extremeness of an observation in the multivariate or high-dimensional space.  We calculate the Mahalanobis distance for the influential point as compared to the original distribution of multivariate normal points (without regard to the different group labels).

@@

## Key Questions to Answer

Is there a way to look at the individual contribution of each feature to the influential points? (simulate this?)

>>>>

---
title: "Formal Paper Draft"
author: "Paul Harmon"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  pdf_document: default
  html_document: default
subtitle: Influence Diagnostics for High-Dimensional Ordination Techniques
---

@@

$$Cost = \sum_{i} \sum_{j} p_{j|i} log(\frac{p_{j|i}}{q_{j|i}})$$

t-SNE is an improvement over Stochastic Neighbor Embedding (SNE), which makes use of Gaussian distributions to model both $p_{ij}$ and $q_{ij}$.  The use of a light-tailed distribution for the lower dimension leads to a phenomenon known as the "crowding problem" where the 2-dimensional map cannot accommodate many neighbors in high dimensions (Van der Maaten, 2011). By utilizing a t-distribution for the $q_{ij}$, t-SNE better preserves the local structure of the data. Most of the minimization of objective functions is done via gradient descent (and some methods require simulated annealing because the algorithm can get caught in local optima). Critically, because the optimization is non-convex, and as a consequence of the gradient descent methods used to optimize it, t-SNE has the potential to provide different maps when run on the same data at the same perplexity.

@@

Additionally, t-SNE is subject to a pre-processing step involving principal component analysis (PCA).  In most implementations, "whitening" via PCA is performed on the initial dataset (or input distance matrix) to reduce the dimensionality to a reasonable value.  For instance, in Van der Maaten's initial paper, several examples reduced the initial dimensionality of the data to 30 for computational gains (Van der maaten, 2011).

@@

Campell, Caudle and Hoover (2019) explored different methods than PCA as an intermediate dimensionality reduction step in t-SNE. They compared PCA, Independent Component Analysis (ICA), Linear Discriminant Analysis (LDA), Sammon Mapping, and Locally Linear Embedding (LLA) as preliminary steps in t-SNE and found that resulting maps tended to form different groupings with variation in separation.  While we did not consider assessing the impact of different intermediate dimensionality reduction techniques in this analysis, it does present an interesting route for additional research.

@@

While the focus of this paper is on t-SNE and MDS, there are manifold additional methods for generating lower-dimensional representations of high-dimensional data.  We include some additional methods that were either compared directly with t-SNE in Van der Maaten and Hinton's original paper (2008) or because they represent the cutting edge of high-dimensional data visualiztion tools.  They range in terms of complexity, with some adding relatively minor changes to the MDS structure to those that share similarities with other branches of mathematics, such as topology or topoligical data analysis.  Sammon Mapping is a non-linear version of classical scaling which scales the stress metric by the pairwise distances, attempting to better maintain local structure by equalizing the impact of large and small distances (Sammon, 1969).  The objective function for Sammon mapping is as follows, where $y_i$ and $y_j$ are the ith and jth points in the lower-dimensional representation and $d_{ij}$ represents the Euclidean distance between points $x_i$ and $x_j$ in the high-dimensional space:

$$\phi(Y) = \frac{1}{\sum_{ij}d_{ij}}\sum_{ij}\frac{(d_{ij}^2 - ||y_i-y_j||^2)}{d_{ij}}$$
Isomap similarly extends MDS to be more flexible, and is a non-linear reduction technique (Tenenbaum et al., 2000). It was initially used as a comparison for t-SNE by van der Maaten (2011) and is utilized as a comparison mapping technique here as well.

Stochastic Neighbor Embedding, or SNE, was a precursor to t-SNE and relies on gaussian distributions rather than the t-distrubution specified previously.  It is subject to a problem known as "crowding" which led to the development of t-SNE. It is worth noting that SNE need not be based on symmetricized KL-divergences, as t-SNE is.

Finally, UMAP is a modern competitor for t-SNE that is based topological data science techniques; chiefly, a model of a Riemannian manifold used to approximate the data (McInnes and Healy, 2018).

@@

Belsley, Kuh and Welsch define influential points as a data point “which, either individually or together with several other observations, has demonstrably larger impact on the calculated values of various estimates (coefficients, standard errors, t-values, etc.) than is the case for most of the other observations.”(1980 page 11).  In regression modeling, as in most statistical methods, care must be taken to properly distinguish between an outlier and a so-called influential point.  An outlier is typically denoted as a point that is either much larger or smaller than similar observations on a specific scale. However, an outlier may not necessarily be influential; a truly influential point must be outlying enough relative to similar observations that it impacts the fit of the regression line. The identification of influential points that can impact the results of a model is commonplace and has been the subject of substantial research for regression-based methods.

@@

Although Cook's distance is essentially an F statistic from an $F_{p, n-p}$ distribution, and the method does not use p-values to determine how "influential" a point is. Rather, the F-statistics themselves are typically compared to cutoff values, or observations with relatively large Cook's distance values are considered for investigation of influence.

## Identifying Outliers in Maps

@@

Alternatively, archetypal analysis (Cutler & Breiman, 1994) can be used to identify outlying points in a multivariate space.  The method approximates a convex hull that lies around the boundary of a dataset based on "archetype" points; these extreme points have been used to identify potential multivariate outliers (Catherine's PhD dissertation).  While this method may be effective at identifying and categorizing extreme points on the boundary of the multivariate space, it does not necessarily answer the question of how the inclusion/exclusion of those points might change the shape of a lower-dimensional representation.  This is the problem we seek to answer with our method.

## Permutational Multivariate Analysis of Variance

Although the previous methods give rise to identifying extreme points, they do not clearly inform the influence of a given observation on the resulting shape of the ordination.  Since the goal of this research is to identify a method for identifying and quantifying influential observations in ordination problems, our goal is to extend the ideas that underlie Cook's Distance from a regression framework to a multivariate response framework.  This necessitates the use of distance matrices rather than raw observations, and requires the use of multivariate Analysis of Variance methods to generate something similar to a F-statistic.

Permutational Multivariate Analysis of Variance (PERMANOVA) is a tool for allows for partitioning variation across multivariate data (Anderson, 2017) in other words, it allows for generalization of ANOVA-type tools on multivariate datasets.  Much like some of the multivariate extensions of ANOVA, PERMANOVA allows for testing of differences in means while taking into account the correlation structures between variables.

A critical difference between some of the standard tools for performing MANOVA and PERMANOVA is that the latter method does makes no assumption of multivariate normality. It only requires an assumption of exchangability between  Instead, PERMANOVA relies on the use of permutation tests to generate a pseudo-F test statistic. In a similar vein as ANOVA, PERMANOVA requires the calculation of among-group sum of squares (SSA) and residual within-group sum of squares (SSW).  The total sum of squares is calculated as follows:

$$SS_T =  \sum_{i = 1}^{N-1} \sum_{j = i+1}^{N} \frac{d_{ij}^2}{N}$$ 
Partitioning can be obtained via subtraction of the residual sum of squares $SS_W$ from the total sum of squares (as in ANOVA), which is defined as the sum of each of the groups' squared distances between the observed data and the group centroid, as defined below for the $l^{th}$ group and $g$ is total of number of groups present in the data:

$$SS_w = \sum_{l = 1}^{g} W_l$$ 
The $W_l$ is thus defined as follows, with $n_l$ defined as the the group sample size (such that $\sum_{l=1}^{g}l_g = N$) and where $\epsilon^{[l]}$ is a binary indicator between the ith and jth samples that equals 1 when samples are in the lth group and 0 otherwise:

$$W_l = \sum_{i = 1}^{N-1} \sum_{j =(i +1)}^{N}\epsilon_{ij}^{l} \frac{d_{ij}^2}{n_l} $$

The resulting partitioning of $SS_T - SS_W$ allows for calculation of a pseudo F-test statistic (which is a multivariate analogue for a standard F ratio).  This tests the null hypothesis that there are no differences in the positions of the group centroids in the distance matrices is as follows, where N is the number of observations and g is the number of columns in the design matrix:

$$F = \frac{SS_A}{SS_R}*(\frac{N-g}{g-1}) $$

P-values are then obtained in a typical permutation-based method, meaning that they are directly calculated based on the number of observed permutations that are more extreme than a given result.  For the purposes of this paper, all PERMANOVA models were performed using the R package 'vegan' (Oskanen, et al).

@@

Second, distance matrices contain information on point-to-point distances (or dissimilarities) and, when vectorized, the correlation of the distance matrices measures similarity of two configurations Mantel's test (Mantel, 1967) makes use of this approach to assess correlation between distance matrices; in this case, we simply compare the similarities in the different resulting map distances.

Third, similarities (or correlations) can be converted to distances; one such metric for doing this is $\sqrt{2(1-r_{ij})}$ (James et al, 2013) where $r_{ij}$ is the correlation between $x_i$ and $x_j$.

@@

Then, we utilize PERMANOVA to assess the statistical significance of the differences in our resulting distance matrices. Note that while Anderson (2017) points out PERMANOVA is less robust than its competitors to choice of dissimilarity, this is not of importance for our method as we utilize only Euclidean distance measures in this step.  Our PERMANOVA-based approach is as follows:

@@

Correcting p-values for multiple testing may or may not be necessary depending on how the p-values are intended to be used, although we recommend using Benjamini-Hochberg False Discovery Rate over more conservative Bonferroni corrections (or something similar, assuming that the results are based on p-values as in the Bonferroni Outlier test (Cook & Weisberg, 1982)). We would expect that if the maps are suitably distorted, or *influenced* by a single point, the resulting p-value for the jth PERMANOVA test would be small if the sample sizes are large enough, and quite often will be the smallest p-value generated.

## Generation of P-Values

In a typical PERMANOVA model, the number of permutations that could be generated from a covariate could be quite large, given that in most cases the majority of the values of the covariate may be nonzero.  However, because of the structure of the method presented here, the predictor variable is always a vector with a single indicator 1 in the holdout index and 0's in all of the remaining indices. Thus, there are only n possible permutations for the predictor variable in the PERMANOVA model.

The p-value for each test is then directly calculated as the number of observations that could be more extreme than a given observation, meaning that a lower bound on the p-value is $1/n$; roughly, the set of obtainable p-values would then be directly calculated as $\frac{k}{n}$ where $k$, an integer value between 1 and $n$, refers to the number of cases that are more extreme than the observed result.

In some cases, it may make sense to consider a multiple testing correction for the p-values generated, considering that there are $n$ tests being performed (one for each holdout).  In large data settings, it may be feasible to use such methods to control for the rate of spurious detections; however, in small data settings this may be impractical.  Indeed, in a very small data setting, assuming a typical $\alpha = 0.05$, if $n< 20$, it would be impossible to obtain a p-value of significance at all!

The p-values can instead be construed as providing a sorting of the observations based on their relative extreme-ness.  It stands to reason that, much in the way that Cook's Distance uses F-statistics to identify abberant observations, the pseudo-F statistic or sorted p-value can be compared based solely on relative magnitude. In a setting where only a single influential point is present, we would expect to see that influential point have the smallest p-value (or, equivalently, the largest pseudo-F statistic) of those obtained.

@@

@@

```{r, echo = FALSE}
#2021-10-18_simulationsmaller_mds_list

@@

#mean detection rates
Category <- c("Correctly Identified Influential", "Correctly Non-Identified Non-Influential")
TSNE <- c(round(mean(ci1),3), round(mean(cni1),3))
MDS <- c(round(mean(ci2),3), round(mean(cni2),3))
Isomap <- c("","")
Sammon <- c("","")
temp <-rbind(Category,TSNE, MDS, Isomap, Sammon)
colnames(temp) <- Category
kable(temp)

@@

#mean detection rates
Category <- c("Correctly Identified Influential", "Correctly Non-Identified Non-Influential")
TSNE <- c(round(mean(ci1),3), round(mean(cni1),3))
MDS <- c(round(mean(ci2),3), round(mean(cni2),3))
Isomap <- c("","")
Sammon <- c("","")
temp <-rbind(TSNE, MDS, Isomap, Sammon)
colnames(temp) <- Category
kable(temp)

@@

#mean detection rates
#mean detection rates
Category <- c("Correctly Identified Influential", "Correctly Non-Identified Non-Influential")
TSNE <- c(round(mean(ci1),3), round(mean(cni1),3))
MDS <- c(round(mean(ci2),3), round(mean(cni2),3))
Isomap <- c("","")
Sammon <- c("","")
temp <-rbind(Category,TSNE, MDS, Isomap, Sammon)
colnames(temp) <- Category
kable(temp)

@@

In order to determine a notion of how severe an outlier/influential point might be, we investigated several methods for quantifying the extremeness of an observation in the multivariate or high-dimensional space.  We calculate the Mahalanobis distance (Mahalanobis, 1936)  for the influential point as compared to the original distribution of multivariate normal points (without regard to the different group labels).

Given a vector $\vec{x}$ with p elements,the Mahalanobis distance is calculated as follows, given mean $mu$ and covariance matrix $\Sigma$:

$$D^2 = (\vec{x} - \mu)^t \Sigma^{-1} (\vec{x} - \mu) $$
The mahalanobis distance can be calculated across each of the simulated datasets and compared to get a sense for how the efficacy of the PERMANOVA-based method changes when influential points are only mildly outlying compared to very extreme points. *Ideally, we would see detection rates that are substantively larger when assessing influential points that have a larger mahalanobis distance than those that have a smaller mahalanobis distance.  This allows for us to assess a power curve to determine how the method is able to detect differences of different magnitudes. *

```{r}
#show plot with number of features on X axis and detection rate on y axis

plot(c(1,2,3,4,5), c(.5,.6,.7,.9,.9), xlab = "# Features", ylab = "Detection Rate", type = "l", main = "Placeholder For Power Curve")
```

## Sparsity of Influential Features

The number of features on which an observation is anomolous is potentially an additional critical element, and can impact how the method detects influence.  Per the previous simulation set up, we simulated observations from the central group with means that differed on a single feature.  We then increased the number of features that differed from the central group mean up to $p-1 = 9$ features to see how the method would detect differences based on influential points that are only influential on a small number of covariates.

```{r}
#show plot with number of features on X axis and detection rate on y axis

df3list <- readRDS("C:/Users/paulh/Documents/Doctoral Work/InfluenceInMaps/2021-11-21_SparsitySimulation_mds_list.RDS")

TF <- matrix(TRUE, nrow = length(df3list[[1]]), ncol = length(df3list))
templist <- list()

for (i in 1:(length(df3list)-1)){

  temp <- rep(TRUE, 100)
  ## Identification Rate at given K
  for(j in 1:length(df3list[[1]])){
    temp[j] <- tail(df3list[[i]][[j]]$PValues,1)<0.05

  }
  TF[,i] <- temp
  templist[[i]] <- sum(temp)/100

}

plot(x = c(1,2,3,4), y = unlist(templist), type = "l", xlab = "Number of Points In Different Mean Group", ylab = "Detection Rate")
points(x = c(1,2,3,4), y = unlist(templist), pch = 20)
title("Detection Rates By Sparsity of Outlier - MDS")
```

In truly high dimensional data, it is possible that an observation may only differ substantively on a small number (say 1 or 2) of features in a dataset with potentially thousands of dimensions.  The computational cost of PERMANOVA on such a large dataset would be expensive in such a case.

@@

## Future Work

There are several considerations for future work that might be worth considering.  Based on the differences in t-SNE maps produced by Campell, Caudle and Hoover (2019), an exploration of the impact of different intermediate dimensionality reduction step in t-SNE could be interesting.  It is certainly possible that using PCA as an intermediate method for reducing the data dimensionality prior to t-SNE, or choosing a different dimension reduction technique, could either enhance or hamper the ability of this PERMANOVA-based method to identify influential points.

For several of these methods, we initially examined metrics that assess the individual contribution of each feature to the cost function for specific mapping techniques.  Interestingly, for t-SNE, the individual costs contributions for the points that were simulated as influential were relatively small, and observations that had high individual cost were often not obviously outlying in the maps created.  Some additional research into the dynamics of individual cost contributions - and how tightly these tie to the resulting changes in ordinations when points are included/excluded from maps - could provide an interesting avenue for future work.

Additionally, examining additional mapping techniques may be worthwhile as well. While this paper assesses a linear method for doing dimension reduction (Classical MDS) and a non-linear method (t-SNE), there exists a bevy of other methods that might be worth examining, including UMAP, Sammon Mapping, Isomap, and others.
